Prediction of Protein-Protein Interactions on the Human and Rice Interactome

### 2020-08-21 nicolas/PPIPL3/ReportV2.tex

%% LyX 2.3.5-1 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[onecolumn,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage{todonotes}

@@

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and
Rice Interactome}
\author{Nicol\'as Antonio L\'opez Rozo}
\maketitle
\begin{abstract}
Previous Network-based efforts to predict unmapped protein-protein interactions (PPI's) suggest that proteins with multiple paths of length~3 (L3) are more likely to be connected. This paper extends this so-called L3 principle by taking into account feature extraction and using \texttt{XGBoost} techniques for prediction. In particular, we train the model using handcrafted features as well as features learned from embeddings using \texttt{Node2Vec}. Our main result shows that while L3 remains an important principle for predicting links, the approach is outperform by using embedded features. The mentioned approaches are compared using the human and the rice interactomes.

% need include a feature that compares the dimension of the embedding is useful for prediction with L3 to show that they independent measures - we do not want to be learning L3 through node2vec
\end{abstract}

\section{Introduction}
% need to add other relevant references 
Proteins, as a fundamental constituents of any living being and their
``building blocks'' in terms of biological functionality, interact
among themselves and with other molecules according to their amino
acid sequences and specific three-dimensional structure and folding.
These interactions also depend on many physical and chemical properties
of each macro-molecule and on the cellular organelle where they operate,
such as pH and temperature. For this reason, experimental ways to
reproduce all of the necessary conditions to validate if an interaction
actually occurs within the cell are generally expensive.

The most common way to validate PPI's is the \emph{Yeast-Two-Hybrid}
technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on the expression of a specific reporter gene that
activates by the binding of a DNA-binding Domain (DB) and an Activation
Domain (AD) of a Transcription Factor. For the Y2H technique, a protein
is fused to the DB domain (known as \emph{bait}) and another one to
the AD (known as \emph{prey}). If the proteins do not interact, then
the reporter gene is not expressed. Otherwise, the reporter gene expression
is activated by the activation domain.

@@

Having several of these Y2H results allows scientists to establish
a PPI network, where all known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
unknown interactions. In this report, three prediction methods are
presented and their results are shown: Common Neighbors (\textbf{CN}),
which uses the length-2 path count; length-3 path count (\textbf{A3});
and degree-normalized length-3 paths score (\textbf{L3}).

\begin{figure}[h]
\caption{Protein-Protein Interaction Prediction Scheme}

\noindent \centering{}\includegraphics[width=0.9\columnwidth]{PPI_prediction}
\end{figure}

Focus of the present study is to evaluate different methods for predicting
protein-protein interactions (PPIs) using the existing knowledge of
the network, which is an undirected graph. The traditional way is
usually based on social networks analysis, more specifically on the
Triadic Closure Principle (TCP), that states that the more common
shared friends that two people have, the more likely that they know
each other. As shown in the next sections, the mentioned approach
fails because it does not consider the structural and chemical properties
of the proteins.

For achieving the described results, several networks are used and
compared using state-of-the-art methods, as well as the proposed ones
(CN, A3, L3). For this report, the human interactome was used (\emph{HI-II-14}).
A curated version of the mentioned network (\emph{HI-TESTED})
was also used in order to compare the influence of research bias on
the network predictions. A massive experimental assay was carried
on and its results are consolidated and used to build a validation
network (\emph{HI-III}).

@@

Human interactome data and base source code were downloaded from the
repository of the length-3 degree normalized paths methodology~\cite{Kovacs2019}:
the dataset \emph{HI-II-14} and \emph{HI-TESTED} are used
for prediction and the dataset \emph{HI-III} is used for validation.
Rice interactome information was downloaded from the STRING database~\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
The downloaded file was \emph{4530.protein.links.detailed.v11.0.txt}.
and contains more than 8 million PPIs from several resources. For
the purpose of this study and based on the previous work, only PPIs
with evidence from curated databases were used (rows where \emph{databases}
column has a value greater than zero), resulting in a network with
NNNN nodes and MMMM edges.

@@

Previous code implementation was adapted from \texttt{C++} to \texttt{Python}
(3.6), in order to unify the algorithms into one single script. For
the purpose of algorithmic validation, the three methods were implemented
from scratch with basic functionalities and data structures of the
\emph{Python} (V3.6) language.

@@

Information for the human interactome was used as-is, which corresponds
to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
$9433$ and $23322$ interactions. For the rice interactome, an additional
preprocessing was performed. The filtered network for rice consists
of $5025$ proteins (nodes) and $164420$ interactions (edges) distributed
among $178$ connected components. The connected component with the
greatest number of edges was selected in this case.

The extracted connected component consists of $n=4390$ nodes and
$m=163319$ edges, which corresponds to $99.33%\%
$ of filtered edges. Further investigation is applied to this network,
which is very similar in number of nodes to the curated information
on the human interactome, although rice network is much more connected.

@@

For the interaction prediction for each network, the algorithms described
below were used. It is important to keep in mind how the protein-protein
interaction (PPI) network $G=(V,E)$ is conceptualized: each node
($v_{i}\in V$) represents a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
represents and interaction among proteins $v_{i}$ and $v_{j}$.

\begin{figure}[h]
\caption{Edge Prediction Paradigms}

\noindent \centering{}\includegraphics[width=0.8\columnwidth]{TCP_extract.png}
\end{figure}

\begin{description}
\item [{Common~Neighbors~(CN)}] This method is based on the Triadic Closure
Principle: ``the more common friends two individuals have, the more
likely that they know each other''. For the implementation of this
method, $A{{}^2}$ matrix is calculated, being $A$ the adjacency
matrix of the network.
\item [{Length-3~Paths~(A3)}] This is the simplest implementation of
the proposed insight of ``if my friends and your friends interact,
then we might interact too''. The calculating is carried on with
$A{{}^3}$, i.e, the third power of the adjacency matrix.
\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
overestimate the importance of some edges due to intermediate hubs
which add many shortcuts in the graph. To address that issue, a degree
normalization for the path $X\rightarrow U\rightarrow V\rightarrow Y$
is applied by considering the degree $k$ of the intermediate nodes
$U$ and $V$, as follows.
\[
p_{XY}=\sum_{U,V}\frac{A_{XU}\cdot A_{UV}\cdot A_{VY}}{\sqrt{k_{U}\cdot k_{V}}}
\]
\\
where $A_{ij}$ represents the value of the adjacency matrix for nodes
$i$ and $j$: 1 if the edge $\{i,j\}$ exists, 0 otherwise.
\end{description}

\subsection{Sampling and Precision Counting}

Sampling proceeds as follows: first, a fraction (10\%) of the edges
is removed at random from the network. Then, for each method (\textbf{CN},
\textbf{A3}, \textbf{L3}) the prediction of the removed fraction is
performed. The new edges are then compared against the original (removed)
edges according to the ranking that each method yields. Edges belonging
to the original connected component are considered positive in the
sense that they are experimentally validated interactions. For each
of the tested networks and a rank $r$, the precision was calculated
as 
\[
P_{r}=\frac{L_{r}}{r}
\]
\\
where $L_{r}$ is the total number of positive edges in the range
$[1,r]$.

The previous procedure was applied for a fraction of 0.90 remaining
edges, and a mean of 10 simulations was obtained.

@@

The selected parameters for the model were \texttt{max\_depth=3}, \texttt{colsample\_bytree=0.6}
and \texttt{eval\_metric=`auc'}.
\subsection{Result Validation}

For applying the test results, a partition of the whole dataset was
made: 80\% of the data was randomly selected and used for training,
while the remaining 20\% was used for validation. The whole training-validation
procedure was applied 5 times.

@@

\section{Results}

As explained above, the precision trend for each of the networks are
presented for the predicted top 2000 interactions.

@@

\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-ii-14.png}
\end{figure}

@@

\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-tested.png}
\end{figure}

@@

Previous result comes as no surprise when the biological basis of
protein interactions is considered: It is necessary that protein A
and protein B have complementary structures in order to interact,
and when classical paths of length~2 are used, the predicted protein
interactions usually have the same structures, and not complementary
ones.

% please update the conclusions - WSD does not seem relevant 
For the analyzed networks, it is also important to highlight their
robustness when compared to a random network with the same density:
if weighted spectral distribution (WSD) is used, then the used networks
are on average twice as robust as their random network versions.

\bibliographystyle{plain}
\bibliography{refs.bib}

>>>>

%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{verbatim}
\usepackage{float}
\usepackage{graphicx}

@@

\usepackage{babel}
\begin{document}
\title{Prediction of Protein-Protein Interactions on the Human and Rice Interactome }
\author{Nicolás Antonio López Rozo}
\maketitle
\begin{abstract}

Previous Network-based efforts to predict unmapped protein-protein
interactions (PPI's) suggest that proteins with multiple paths of
length 3 (L3) are more likely to be connected. This paper extends
this so-called L3 principle by taking into account feature extraction
and using \texttt{XGBoost} techniques for prediction. In particular,
we train the model using handcrafted features as well as features
learned from embeddings using \texttt{Node2Vec}. Our main result shows
that while L3 remains an important principle for predicting links,
the approach is outperform by using embedded features. The mentioned
approaches are compared using the human and the rice interactomes.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

\section{Introduction}

\begin{comment}
Need to add other relevant references
\end{comment}

Proteins are the chief actors of biological functionality inside of
the cell, and as such, do not usually work as single substances but
rather as a part of dynamic networks, interacting with other substances
for a wide variety of purposes \cite{Lin2017}.

Protein-protein interactions (PPIs) have a key role in a variety of
biological processes such as signal transduction, homeostasis control,
stress responses, plant defense and organ formation. On the molecular
level, PPIs have importance in protein phosphorylation, transcriptional
co-factor recruitment, transporter activation and many others, thus
playing an essential role in many physiological and developmental
processes in virtually all organisms \cite{Zhang_2010_PPI}.

Prediction of potentially relevant, yet unexplored PPIs is a current
research topic on bioinformatics and therefore, several authors have
proposed different methods for extrapolating information from the
existing PPI networks. As presented by Kovacs et al (2019), information
related to counting paths of length 3 (L3) seems to outperform state-of-the-art
methods when it comes to predict interactions among proteins for a
variety of model organisms such as yeast (\emph{S. cerevisiae}), Arabidopsis
(\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
fission yeast (\emph{S. pombe}) and mouse (\emph{M. musculus}), as
well as on the human interactome \cite{Kovacs2019}.

The most common way to validate PPI's is the \emph{Yeast-Two-Hybrid}
technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on the expression of a specific reporter gene that
activates by the binding of a DNA-binding Domain (DB) and an Activation
Domain (AD) of a Transcription Factor that binds to an Upstream Activation
Sequence (UAS). For the Y2H technique, a protein is fused to the DB
domain (known as \emph{bait}) and another one to the AD (known as
\emph{prey}). If the proteins do not interact, then the reporter gene
is not expressed. Otherwise, the reporter gene expression is activated
by the activation domain.

\begin{comment}
NL: I will make this figure myself
\end{comment}

@@

Having several of these Y2H results allows scientists to establish
a PPI network, where all known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
unknown interactions. In this report, three prediction methods are
presented and their results are shown: Common Neighbors (\textbf{CN}),
which uses the length-2 path count; raw count of paths of length 3
(\textbf{A3}); and degree-normalized length-3 paths score (\textbf{L3}).

Focus of the present study is to evaluate different methods for predicting
protein-protein interactions (PPIs) using the existing knowledge of
the network, which is an undirected graph. The traditional way is
usually based on social networks analysis, more specifically on the
Triadic Closure Principle (TCP), that states that the more common
shared friends that two people have, the more likely that they know
each other. As shown by previous studies, the mentioned approach fails
because it does not consider the structural and chemical properties
of the proteins \cite{Kovacs2019}.

For achieving the described results, human and rice PPI networks are
used and compared using state-of-the-art methods, as well as the proposed
ones (CN, A3, L3). In the case of the human network, human interactome
(\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED})
were used. A massive experimental assay was carried on and its results
were consolidated and used to build a validation network (\emph{HI-III}).

@@

Human interactome data and base source code were downloaded from the
repository of the length-3 degree normalized paths methodology \cite{Kovacs2019}:
the dataset \emph{HI-II-14} and \emph{HI-TESTED} are used for prediction
and the dataset \emph{HI-III} is used for validation.

Rice interactome information was downloaded from the STRING database
\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
The downloaded file was \emph{4530.protein.links.detailed.v11.0.txt}.
and contains more than 8 million PPIs from several resources. For
the purpose of this study and based on previous work, only PPIs with
evidence from curated databases were used (i.e. rows where the column
\emph{databases} has a value greater than zero), resulting in a network
with $5025$ nodes and $164420$ edges.

@@

Previous code implementation was adapted from \texttt{C++} to \texttt{Python}
(V3.6), in order to unify the algorithms into one single script. For
the purpose of algorithmic validation, the three methods were implemented
from scratch with basic functionalities and data structures of the
\texttt{Python} language.

@@

Information for the human interactome was used as-is, which corresponds
to networks of $4298$, $3727$ and $5604$ proteins and $13868$,
$9433$ and $23322$ interactions.

For the rice interactome, an additional preprocessing was performed.
The filtered network for rice consists of $5025$ proteins (nodes)
and $164420$ interactions (edges) distributed among $178$ connected
components. The connected component with the greatest number of edges
was selected in this case. The extracted connected component consists
of $n=4390$ nodes and $m=163319$ edges, which corresponds to $99.33%\%
$ of filtered edges. Further investigation is applied to this network,
which is very similar in number of nodes to the curated information
on the human interactome, although rice network is much more connected.

@@

For the interaction prediction for each network, the algorithms described
below were used. It is important to keep in mind how the protein-protein
interaction (PPI) network $G=(V,E)$ is conceptualized: each node
($v_{i}\in V$) represents a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
represents and interaction among proteins $v_{i}$ and $v_{j}$.
\begin{description}
\item [{Common~Neighbors~(CN)}] This method is based on the Triadic Closure
Principle: ``the more common friends two individuals have, the more
likely that they know each other''. For the implementation of this
method, $A{{}^2}$ matrix is calculated, being $A$ the adjacency
matrix of the network.
\item [{Length-3~Paths~(A3)}] This is the simplest implementation of
the proposed insight of ``if my friends and your friends interact,
then we might interact too''. The calculating is carried on with
$A{{}^3}$, i.e, the third power of the adjacency matrix.
\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
overestimate the importance of some edges due to intermediate hubs
which add many shortcuts in the graph. To address that issue, a degree
normalization for the path $X\rightarrow U\rightarrow V\rightarrow Y$
is applied by considering the degree $k$ of the intermediate nodes
$U$ and $V$, as follows.
\[
p_{XY}=\sum_{U,V}\frac{A_{XU}\cdot A_{UV}\cdot A_{VY}}{\sqrt{k_{U}\cdot k_{V}}}
\]
\\
where $A_{ij}$ represents the value of the adjacency matrix for nodes
$i$ and $j$: 1 if the edge $\{i,j\}$ exists, 0 otherwise.
\end{description}

\subsection{Sampling Procedure}

For each network of protein interactions, the following procedure
was performed 10 times in order to address the stochastic nature of
the process and have a consensus:
\begin{itemize}
\item A percentage of interactions is removed at random from the network
(20\%).
\item The same amount of removed interactions are then predicted using the
main methods for prediction mentioned by Kovacs et al (2019): Common
Neighbors (\textbf{A2}), raw path count of paths of length 3 (\textbf{A3})
and the Length-3 degree-normalized score (\textbf{L3}).
\item A test dataset is created as follows: all removed edges are included
(as observed positives for the ML algorithm) and from the predicted
edges of \textbf{A2}, \textbf{A3} and \textbf{L3} that don't lie in
the previous classification (observed negatives), a random subset
is chosen such that the dataset is balanced, that is, the amount of
observed positive labels is equal to the observed negative labels.
\item Once the dataset is ready, it is randomly partitioned: 80\% is used
for \texttt{XGBoost} model training and 20\% is used for validation.
It is important to have in mind that balanced distribution of the
positive and negative labels in the datasets was satisfied.
\end{itemize}

@@

The selected parameters for the model were:\texttt{ max\_depth=3},
\texttt{colsample\_bytree=0.6} and \texttt{eval\_metric=''auc''}.

\subsection{Result Validation}

As mentioned before, 80\% of the final dataset was randomly selected
and used for training, while the remaining 20\% was used for validation.
The whole training-validation procedure was applied 10 times.

@@

\section{Results and Discussion}

\subsection{Rice Interactome}

For the rice interactome, the different model-features combinations
were trained and validated. After executing the mentioned routines,
the results are shown in Figure \ref{F1}. First, one should have
a baseline of comparison, which in this case corresponds to \texttt{Node2Vec}
without any additional feature included. The plot below shows those
results, and one can see that its mean performance using the AUC metric
is 0.52, and that the results among the 10 repetitions are consistent.
This results mean that the model using only the default features perform
barely as good as a random choice of the labels.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F1}Summary ROC curves for \texttt{Node2Vec} model alone}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.95\columnwidth]{Only_ML/ROConlyML_SUMMARY}
\end{figure}

\section{(PENDING FROM HERE ON)}

@@

\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-ii-14.txt\string".eps}
\end{figure}

@@

\noindent \centering{}\includegraphics[width=1\columnwidth]{\string"/home/ocin/Documents/02 Asignaturas/10 GraphAnalytics/Project1/hi-tested.txt\string".eps}
\end{figure}

@@

Previous result comes as no surprise when the biological basis of
protein interactions is considered: It is necessary that protein A
and protein B have complementary structures in order to interact,
and when classical paths of length 2 are used, the predicted protein
interactions usually have the same structures, and not complementary
ones.

\begin{comment}
please update the conclusions - WSD does not seem relevant
\end{comment}

For the analyzed networks, it is also important to highlight their
robustness when compared to a random network with the same density:
if weighted spectral distribution (WSD) is used, then the used networks
are on average twice as robust as their random network versions.

\bibliographystyle{plain}
\bibliography{refs}
